---
title: "BST249 HW6"
author: "Sijia Huo"
date: "4/25/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### 1.

```{r}

# initial distribution (m): pi
# transition matrix (m * m): T
# emission matrix (m * k): phi
# observed sequence (n): x

# helper function for log-sum-exp
# take log of elements to sum as input

log_sum_exp = function(log_vec){
  max = max(log_vec)
  log_sum = max + log(sum(exp(log_vec-max)))
  return(log_sum)
}

# forward function
forward = function(log_pi,log_T,log_phi,x){
  
  # construct matrix to hold result
  n = length(x)
  k = ncol(log_phi)
  m = nrow(log_phi)
  log_s_matrix = matrix(0,n,m) # x by z
  
  # dynamic programming
  
  log_s_matrix[1,] = log_pi + log_phi[,x[1]] # for x1

  for(j in 2:n){
    for(i in 1:m){
      # log-sum-exp
      log_s_matrix[j,i] = log_sum_exp(log_s_matrix[j-1,] + log_T[,i] + log_phi[i,x[j]])
    }
  }

  log_px = log_sum_exp(log_s_matrix[n,])
  return(list(mat = log_s_matrix, px = log_px))
}

# backward
backward = function(log_T,log_phi,x){
  
  # construct matrix to hold result
  n = length(x)
  k = ncol(log_phi)
  m = nrow(log_phi)
  log_r_matrix = matrix(0,n,m) # x by z
  
  # dynamic programming

  # when j = n, all entries are 0, no need to change
  for(j in (n-1):1){
    for(i in 1:m){
      # log-sum-exp
      log_r_matrix[j,i] = log_sum_exp(log_T[i,] + log_phi[,x[j+1]] + log_r_matrix[j+1,])
    }
  }
  return(log_r_matrix)
}
```

Check the correctness of the implementation

```{r,tidy=TRUE, tidy.opts=list(width.cutoff=75)}
data = read.table("~/Documents/Spring2022/BST249/Homework/HW6/homework-6-data/x.txt", quote="\"", comment.char="")
data = data[,1]

# forward & backward
forward_ret = forward(log(rep(1/10,10)), log(matrix(1/10,10,10)), log(matrix(1/59,10,59)),data)
backward_ret = backward(log(matrix(1/10,10,10)), log(matrix(1/59,10,59)),data)

# check correctness
print(forward_ret$px) # forward p(x)
print(log_sum_exp(log(rep(1/10,10)) + log(rep(1/59,10)) + backward_ret[1,])) # backword p(x)

```

Based on our setting, we have equal chance observing any of the $k = 59$ words at each of the $n = 292$ time steps. Therefore, we have $p(x_1,\cdots,x_n) = (\frac{1}{59})^{292}$ and $\log p(x_1,\cdots,x_n) = 292\log(\frac{1}{59}) = -1190.641$. Therefore, our implementation is correct.

### 2.

Derive $\varphi_{i}=\left(\varphi_{i 1}, \ldots, \varphi_{i k}\right)$ where $i = 1\cdots m$ using Lagrange multipliers as follows: Given $Q\left(\theta, \theta_{k}\right)=\sum_{i=1}^{m} \gamma_{1 i} \log \pi_{i}+\sum_{t=2}^{n} \sum_{i, j=1}^{m} \beta_{t i j} \log T_{i j}+\sum_{t=1}^{n} \sum_{i=1}^{m} \gamma_{t i} \log f_{\varphi_{i}}\left(x_{t}\right)$ and set $f_{\varphi_{i}}\left(x_{t}\right) = \sum_{K=1}^{k} \mathrm{I}(x_t = K)\varphi_{iK}$, we then have the following equation to estimate $\varphi_{iK}$ taking account of the constraint that $\sum_{K=1}^{k}\varphi_{iK} = 1$:

$$
\begin{aligned}
0 &=\frac{\partial}{\partial \varphi_{iK}}\left(Q\left(\theta, \theta_{k}\right)-\lambda \sum_{K=1}^{k}\varphi_{iK}\right)\\
0 &= \frac{\partial}{\partial \varphi_{iK}}\left( \sum_{t=1}^{n} \sum_{i=1}^{m} \gamma_{t i} \log \left(\sum_{K=1}^{k} \mathrm{I}(x_t = K)\varphi_{iK}\right)\right) - \lambda \\
0 &=  \frac{\partial}{\partial \varphi_{iK}}\left( \sum_{i=1}^{m} \sum_{K=1}^{k} \sum_{t=1}^{n}  \mathrm{I}(x_t = K) \gamma_{t i} \log \left( \varphi_{iK}\right)\right) - \lambda\\
0 &= \frac{\sum_{t=1}^{n}  \mathrm{I}(x_t = K) \gamma_{t i}}{\varphi_{iK}} - \lambda \\
 \varphi_{iK} &= \frac{\sum_{t=1}^{n}  \mathrm{I}(x_t = K) \gamma_{t i}}{\lambda}
\end{aligned}
$$

From the derivation above, we also have $\sum_{t=1}^{n}\mathrm{I}(x_t = K) \gamma_{t i} = \lambda\varphi_{iK}$. Therefore, we further have $\sum_{K=1}^{k}\sum_{t=1}^{n}\mathrm{I}(x_t = K) \gamma_{t i} = \sum_{K=1}^{k} \lambda \varphi_{iK} = \lambda\sum_{K=1}^{k}\varphi_{iK} = \lambda$. Since $\sum_{K=1}^{k}\sum_{t=1}^{n}\mathrm{I}(x_t = K) \gamma_{t i} =\sum_{t=1}^{n}\sum_{K=1}^{k}\mathrm{I}(x_t = K) \gamma_{t i} = \sum_{t=1}^{n} \gamma_{t i}$, we finally have $\lambda = \sum_{t=1}^{n} \gamma_{t i}$ and $\varphi_{iK} = \frac{\sum_{t=1}^{n}  \mathrm{I}(x_t = K) \gamma_{t i}}{\sum_{t=1}^{n} \gamma_{t i}}$


Given $\beta_{t i j}=\mathbb{P}_{\theta_{k}}\left(Z_{t}=i, Z_{t+1}=j \mid x\right)$ (slightly different from the definition on slides) and $p\left(z_{j}, z_{j+1} \mid x_{1: n}\right) \propto p\left(x_{1: j}, z_{j}\right) p\left(z_{j+1} \mid z_{j}\right) p\left(x_{j+1} \mid z_{j+1}\right) p\left(x_{j+2: n} \mid z_{j+1}\right)$, we have 

$$
\begin{aligned}
\beta_{t i j} &= \frac{p\left(x_{1: t}, z_{t}=i\right) p\left(z_{t+1}=j \mid z_{t}=i\right) p\left(x_{t+1} \mid z_{t+1}=j\right) p\left(x_{t+2: n} \mid z_{t+1}=j\right)}{\sum_{i=1}^m\sum_{j=1}^{m}p\left(x_{1: t}, z_{t}=i\right) p\left(z_{t+1}=j \mid z_{t}=i\right) p\left(x_{t+1} \mid z_{t+1}=j\right) p\left(x_{t+2: n} \mid z_{t+1}=j\right)} \\
&= \frac{s_{t}\left(z_{t}=i\right)\cdot T_{ij}\cdot \varphi_{jx_{t+1}} \cdot r_{t+1}\left(z_{t+1}=j\right)}{\sum_{i=1}^m\sum_{j=1}^{m}s_{t}\left(z_{t}=i\right)\cdot T_{ij}\cdot \varphi_{jx_{t+1}} \cdot r_{t+1}\left(z_{t+1}=j\right)}
\end{aligned}
$$

Given $\gamma_{t i}=\mathbb{P}_{\theta_{k}}\left(Z_{t}=i \mid x\right)$ and $p\left(z_{j} \mid x_{1: n}\right) \propto p\left(x_{1: j}, z_{j}\right) p\left(x_{j+1: n} \mid z_{j}\right)$ (from lecture slides), we have $$\gamma_{t i} = \frac{p\left(x_{1: t}, z_{t}=i\right) p\left(x_{t+1: n} \mid z_{t}=i\right)}{\sum_{j=1}^{m} p\left(x_{1: t}, z_{t}=i\right) p\left(x_{t+1: n} \mid z_{t}=i\right)} = \frac{s_{t}\left(z_{t}=i\right)r_{t}\left(z_{t}=i\right)}{\sum_{i=1}^{m}s_{t}\left(z_{t}=i\right)r_{t}\left(z_{t}=i\right)} = \sum_{j=1}^{m}\beta_{tij}$$

With $\pi_{i}=\frac{\gamma_{1 i}}{\sum_{j=1}^{m} \gamma_{1 j}}$ and $T_{i j}=\frac{\sum_{t=1}^{n-1} \beta_{t i j}}{\sum_{t=1}^{n-1} \sum_{j=1}^{m} \beta_{t i j}}=\frac{\sum_{t=1}^{n-1} \beta_{t i j}}{\sum_{t=1}^{n-1} \gamma_{t i}}$ along with $p\left(x_{1: n}\right)=\sum_{z_{n}} s_{n}\left(z_{n}\right)$ (from lecture slides), the implementation of Baum–Welch algorithm is as follows:

```{r,tidy=TRUE, tidy.opts=list(width.cutoff=75)}
library(gtools)

baum_welch = function(x,m,k) {
  
  # random initialization
  log_pi= log(rdirichlet(1,rep(1,m)))
  log_T = log(rdirichlet(m,rep(1,m)))
  log_phi = log(rdirichlet(m,rep(1,k)))
  n = length(x)
  forward_ret = forward(log_pi,log_T,log_phi,x)
  log_px_new = forward_ret$px
  converge = FALSE # flag
  counter = 0 # count the number of ronuds untill convergence
  
  while (!converge){
    
    counter = counter + 1
    
    # E step
    log_s = forward_ret$mat
    log_r = backward(log_T,log_phi,x)
    
    # beta
    log_beta = array(0, dim = c(m,m,n-1))
    for (t in 1:(n-1)) {
      for (i in 1:m) {
        # numerator of beta
        log_beta[i,,t] = log_s[t,i] + log_T[i,] + log_phi[,x[t+1]] + log_r[t+1,]
      }
      log_beta_denom = log_sum_exp(log_beta[,,t])
      log_beta[,,t] = log_beta[,,t]-log_beta_denom
    }

    # gamma
    log_gamma = t(apply(log_beta, c(1, 3), log_sum_exp)) # time by state
    log_gamma_tn = apply(log_beta[,,n-1], 2, log_sum_exp)  # for t=n ; 
    log_gamma = rbind(log_gamma, log_gamma_tn)
  
    # M step
    log_T = apply(log_beta, c(1,2), log_sum_exp) - apply(log_gamma[1:(n-1),], 2, log_sum_exp)
    for (i in 1:m){
      log_phi[i,] = sapply(1:k, function(j) log_sum_exp(log_gamma[which(x == j),i]) - log_sum_exp(log_gamma[,i]))
    }
    
    log_pi = log_gamma[1,]-log_sum_exp(log_gamma[1,])
    
    # check convergence + E step
    forward_ret = forward(log_pi,log_T,log_phi,x)
    log_px_old = log_px_new; 
    log_px_new = forward_ret$px
    converge = (abs(log_px_new - log_px_old) < 0.01)
  }
  
  return(list(log_pi = log_pi,log_T = log_T, log_phi = log_phi, log_px = log_px_new, iter = counter))
}
```

Run the algorithm with different m and initializations.

```{r}
bw_ret = lapply(rep(c(10,30,100),each=3), baum_welch, x = data, k = 59)
```

Print out the number of iterations and the $\log p\left(x_{1}, \ldots, x_{n}\right)$ for each run. For each vector displayed below, the first 3 elements are for $m = 10$, the 4th-6th elements are for $m=30$ and the last three elements are for $m = 100$.

```{r}
# iterations
sapply(bw_ret, function(ret) ret$iter)

# log_p_x
sapply(bw_ret, function(ret) ret$log_px)
```

We notice that the number of iterations are not significantly different across different m and the difference is mainly caused by the random initialization. Nevertheless, as the number of hidden states increase, $p\left(x_{1}, \ldots, x_{n}\right)$ increase. This means that with more hidden states, the overall likelihood at the last step increases and our estimation gets better.

### 3.

```{r,tidy=TRUE, tidy.opts=list(width.cutoff=75)}
# HMM generating function with HMM parameters and sequence length

code <- read.table("~/Documents/Spring2022/BST249/Homework/HW6/homework-6-data/code.txt", quote="\"", comment.char="")
code_indx = code[,2]

seq_gen = function(pi,T,phi,N,word_indx){
  
  # construct matrix to hold result
  k = ncol(phi)
  m = nrow(phi)
  
  # generate hidden states
  z = rep(0,N)
  z[1] = sample(1:m,size = 1,prob = pi) # initial probability
  for(i in 2:N){
    z[i] = sample(1:m,size = 1,prob = T[z[i-1],])
  }
  
  # generate observed x, map back to word
  x = sapply(1:N, function(i) sample(1:k,size=1,prob = phi[z[i],]))
  x_word = sapply(1:N, function(i) word_indx[x[i]])
  x_word = paste(x_word,collapse=" ")
  return(x_word)
}
```

Take the results of the first run for each of $m \in \{10,30,100\}$ in question 2 as the estimated parameters, skip step (a). Generate the estimated sequence (step b and c) as follows.

```{r,echo=FALSE}
source("wrap_long_lines.R")
```

```{r, linewidth=90}
# generate and print out sequence
N = 250

# m = 10
ret_10 = bw_ret[[1]]
print(seq_gen(exp(ret_10$log_pi),exp(ret_10$log_T),exp(ret_10$log_phi),N,code_indx))

# m = 30
ret_30 = bw_ret[[4]]
print(seq_gen(exp(ret_30$log_pi),exp(ret_30$log_T),exp(ret_30$log_phi),N,code_indx))

# m = 100
ret_100 = bw_ret[[7]]
print(seq_gen(exp(ret_100$log_pi),exp(ret_100$log_T),exp(ret_100$log_phi),N,code_indx))

```

Compare the estimated sequence with the original sequence as follows.

```{r, linewidth=90}
# original string
paste(sapply(1:292, function(i) code_indx[data[i]]),collapse=" ")
```

Final comments:

As the number of hidden states increase, our estimation for the latent structure of the text becomes more and more accurate. As a result, the sequence generated using the estimation becomes more and more readable and interpretable. Nevertheless, we still can't restore the original text from our estimation due to the property of the HMM: the hidden state at each time step only depends on the state of the last step. 
